---
title: "data In and Out"
output: html_document
date: "2025-03-20"
---

Let's do some preparation for the class by running the following chunk of codes:

```{r setup, include=FALSE}
#install all the packages for this script
pacman::p_load(
    DBI, # database connection
    odbc, # database connection
    tidyverse,# data cleaning packages
    dbplyr # R package interfacing with SQL db
)
```

# Make Database connection
First, will need to setup ODBC connector object (i.e. Data Source Name, or, DSN).You can also refer to the AWS Service Workbench User Manual for more details. 

Use `DBI:dbConnect` and `odbc::odbc()` function with stored credentials to make database connection. Due to re-direction to Umsystem log-in page with the Shibboleth method, you may not be able to Knit this report at once. 

```{r db_connection}
# you can directly put your credential in the function
myconn <- DBI::dbConnect(
   drv = odbc::odbc(),
   dsn = "snowflake_deid",
   uid = "xsm7f@umsystem.edu",
   pwd = ""  # leave the password empty if using Shibboleth (as you will be prompted to the university log-in portal)
)
```

# Load data in using R function by embedding SQL codes
The classic method for loading data from SQL database is to directly send SQL queries over to the database engine, run the queries and return the result back as a data.frame.

```{r}
dat1a<-dbGetQuery(
    myconn,  # connection object
    "select * from ALS_RILUZOLE_DEATH_FINAL"  # sql codes
)
```

# Load data in using R function interfacing with SQL
There are existing R packages with wrapper function that can interface with SQL. In other words, instead of writing out the SQL codes, you can just call the R function to perform some simple SQL operations, such as read a single table into R and write a single table over to SQL database. 

```{r}
dat1b<-tbl(
    myconn,
    sql("select * from PUBLIC.ALS_RILUZOLE_DEATH_FINAL")
) %>%
  collect()
```

# Load data in using SQL chunk supported by Rmd
R markdown further simplified the above codes for embedding SQL queries by including "SQL code chunks". SQL code chunk is implemented with the chunk header of parameters `sql, connection, output.var`. 

```{sql, connection=myconn, output.var="dat1c"}
select * from ALS_RILUZOLE_DEATH_FINAL
```

# Save intermediate data
Please create a sub-folder under the working directory as `./data` and all raw and intermediate data will be saved their for better management. 

## save dataframe as Rdata file (.rda)
```{r}
saveRDS(dat1a,file="C:/project/data/als_aset.rda")
```

Then, how do we read the .rda file back into R?

```{r}
dat1_rds<-readRDS("C:/project/data/als_aset.rda")
```

## save dataframe as CSV file (.csv)
```{r}
write.csv(dat1a,file="C:/project/data/als_aset.csv")
```

Then, how do we read the .csv file back into R?

```{r}
dat1_csv<-read.csv("C:/project/data/als_aset.csv")
```

## save dataframe as delimiter separated file (.txt)
```{r}
write.table(dat1a,file="C:/project/data/als_aset.txt",sep="|")
```

Then, how do we read the .csv file back into R?

```{r}
dat1_txt<-read.table("C:/project/data/als_aset.txt",sep="|")
```

