---
title: "Confirmatory Data Analysis "
author: "Xing Song"
date: "10/18/2022"
output: html_document
---

Preparation: report setup, load packages, load data

```{r setup}
#load package and data
pacman::p_load(
  tidyverse,
  epitools,
  rcompanion
)
aset<-readRDS("C:/Users/Administrator/Documents/proj/data/final_set.rda")
```

**************************************************************************************************

## Confirmatory Data Analysis (CDA) uses inferential statistics to:

- Evaluate evidence by challenging their assumptions about the data.
- Produce estimates with a specified level of precision
- Test Hypotheses about correlations
- Estimate relationships between variables
- Evaluate the difference between 

**************************************************************************************************


## Categorical Exposure and Categorical Outcome

> Motivating Hypothesis: whether there is a difference of Riluzole use between males and females? 


```{r}
tbl<-table(aset$SEX,aset$RILUZOLE_IND,dnn=list("SEX","Riluzole"))
tbl
```

*Odds Ratio*

Or use existing package, which usually produced more results, such as

*Fisher Exact Test, Mid-p Test, or Chi-Squared test*

```{r}
epitools::oddsratio(tbl)
```

*Phi Correlation Coefficients (Matthews correlation coefficients)*

```{r}
rcompanion::phi(tbl)
```

*Cramer's V statistics*

```{r}
rcompanion::cramerV(tbl)
```


## Categorical Exposure and Numerical Outcome

> Motivating Hypothesis: Whether female has a higher likelihood of longer Riluzole usage?

*two-sample t-statistics*

```{r}
t.test(RILUZOLE_DUR ~ SEX, data = aset)
```

```{r}
t.test(RILUZOLE_DUR ~ SEX, data = aset, alternative="greater")
```

*ANOVA*

```{r}
summary(aov(RILUZOLE_DUR ~ SEX,data=aset))
```

*Kruskal-Wallis Test*

```{r}
kruskal.test(RILUZOLE_DUR ~ SEX,data=aset)
```


## Numerical Exposure and Numerical Outcome

> Motivating Hypothesis: Whether age at first ALS diagnosis is correlated with Riluzole use exposure? 

*Pearson Correlation Coefficient*

```{r}
cor.test(aset$AGE_AT_ALS1DX,aset$RILUZOLE_DUR,method="pearson")
```


*Spearman Correlation Coefficient*

```{r}
cor.test(aset$AGE_AT_ALS1DX,aset$RILUZOLE_DUR,method="spearman")
```

*Kolmogorov-Smirnov Tests*

```{r}
ks.test(aset$AGE_AT_ALS1DX,aset$RILUZOLE_DUR)
```
